Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-04-06 Manuscript/manuscript.Rmd
manuscript edits

While direct overlaps represent the most straightforward measure of functional associations, they do not account for indirect relationships shown to be one of the major driving force of genome regulation [??? REFS]. The most prominent example maybe the enhancer-promoter relationships, with enhancers are frequently located hundreths or thousands bases away from their target promoters [??? REFS]. The use of distances between TAD boundaries and genomic annotations has largely been ignored. However, these distance features have the potential to offer a more accurate spatial representation of which genomic features are associated with TAD boundaries.

@@

By isolating bins that overlapped with flanked TAD boundaries and, in turn, counting the instances in which annotations overlapped with said bins (using the overlap counts discussed above) we were able to assess the correlations of genomic annotations and determine which annotations were enriched together at boundaries.

Clustered correlograms for known colocalized annotations showed that distinct groups of transcription factors were highly correlated including CTCF, ZNF143, RAD21, and SMC3ab (Supplementary Figures 2-9). Likewise, combinations of particular histone modifications and chromatin states were also shown to exhibit similar profiles near TAD boundaries including H3K4me1, H3K4me2, and H3K4me3, aswell as strong and weak enhancers. Furthermore, particular chromatin segmentation states, inclduing heterochromatin, repressed chromatin, poised promotors, and repetitive copy number variations, were found to be highly correlated, yet inversely correlated with many transcription factor binding sites. These results were consistent across resolution and cell line. The high correlations among many of the genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation.

## CLASS IMBALANCE SEVERELY IMPACTS MODEL PERFORMANCE ACROSS RESOLUTIONS

We first evaluated model performances using only the overlap counts predictor types without re-sampling as these characteristics represented models with the least granularity and could act as a baseline. We measured the performances across resolutions for each cell line (Figure 8 (A)). Some notable results became clear. First, we see that the conventional AUC value was unable to accurately portray the poor performance of the models when compared to the composite metrics F1-score, MCC, and, to a lesser extent, the balanced accuracy. Inspecting the F1-score and MCC more closely, we see that some models at lower resolutions (for either cell line) were unable to correctly predict any bins that overlap with flanked TAD boundaries in the test data, as indicated by the missing bars.

>>>>

While direct overlaps represent the most straightforward measure of functional associations, they do not account for indirect relationships shown to be one of the major driving force of genome regulation [??? REFS]. The most prominent example being the enhancer-promoter relationships, where enhancers are frequently located hundreds or thousands of bases away from their target promoters [??? REFS]. The use of distances between TAD boundaries and genomic annotations has largely been ignored. However, these distance features have the potential to offer a more accurate spatial representation of which genomic features are associated with TAD boundaries.

@@

By isolating flanked TAD boundaries and, in turn, counting the instances in which annotations overlapped with boundary regions, we were able to assess the correlations of genomic annotations and determine which annotations were enriched together at boundaries.

Clustered correlograms for known colocalized annotations showed that distinct groups of transcription factors were highly positively correlated in both cell lines including CTCF, ZNF143, RAD21, and SMC3ab (Supplementary Figures 2-9). Likewise, combinations of particular histone modifications and chromatin states were also shown to exhibit similar profiles near TAD boundaries including H3K4me1, H3K4me2, and H3K4me3, as well as strong and weak enhancers. Furthermore, particular chromatin segmentation states, inclduing heterochromatin, repressed chromatin, poised promotors, and repetitive copy number variations, were found to be highly positively inter-correlated, yet were negatively correlated with many transcription factor binding sites. These results were consistent across resolution and cell line. Notably, correlations amoung annotations were found to increase as resolutions decreased for both cell lines. This indicated that TAD bounardies called, and subsequently flanked, at lower resolutions tended to encapsulate more of these epigenomic marks and contribute to higher levels of enrichment.

The high correlations among many of the genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation.

## CLASS IMBALANCE SEVERELY IMPACTS PERFORMANCE OF MODELS WITH OVERLAP COUNTS PREDICTOR TYPES

We first evaluated model performances using only the overlap counts predictor types without re-sampling as these characteristics represented models with the least granularity and could act as a baseline. We measured the performances across resolutions, for each cell line, and found that .

Some notable results became clear. First, we see that the conventional AUC value was unable to accurately portray the poor performance of the models when compared to the composite metrics F1-score, MCC, and, to a lesser extent, the balanced accuracy. Inspecting the F1-score and MCC more closely, we see that some models at lower resolutions (for either cell line) were unable to correctly predict any bins that overlap with flanked TAD boundaries in the test data, as indicated by the missing bars.
